Overlapping Batches for the Assessment of Solution Quality in Stochastic Programs

### 2013-10-25 omrp.tex
Streamlining Sec 4.1, 4.2, and 4.3
a number of edits here and more writing to explain things better.

Taking $n$ observations of the random variables $\xit$ and splitting them into $k$ nonoverlapping batches of size $m$, define 
\begin{align} 
	\gb_j =\  & \frac{1}{m} \sum_{i = 1}^{m} f(\xh, \xit^{m(j-1)+i}) - \min_{\x \in X} \frac{1}{m} \sum_{i=1}^{m} f(\x, \xit^{m(j-1)+i}),& \ \ \ \ j = 1,2, \dots, k \label{eq:mrp_gap}\\
         =\  & \frac{1}{m} \sum_{i = 1}^{m} \left[ f(\xh, \xit^{m(j-1)+i}) -  f(\xs_j, \xit^{m(j-1)+i}) \right],& \ \ \ \ j = 1,2, \dots, k, \notag
\end{align}
where $\xs_j$ denotes an optimal solution to (\ref{eq:sto_prog_m}) of batch $j$ given on the right-hand side of (\ref{eq:mrp_gap}).    
Since we are assessing the quality of a given solution $\xh \in \X$, we suppress it from the notation $\gb_j$.
The optimality gap estimator $\gb_j$ is similar to $\yb_j$ with individual observations $y^{m(j-1)+i} = f(\xh,\xit^{m(j-1)+i}) - f(\xs_j,\xit^{m(j-1)+i})$, for $i=1,\ldots,m$ and $j=1,\ldots, k$.
Notice that, in this case, $y^i$ not only depends on observation $i$ of batch $j$, $\xit^{m(j-1)+i}$, but all the observations in that batch $\xit^{m(j-1)+1},\xit^{m(j-1)+2},\ldots,\xit^{m(j-1)+m}$ because $\xs_j$ is the optimal solution to (\ref{eq:sto_prog_m}) given in (\ref{eq:mrp_gap}).   
As before, after $k$ estimates are obtained, the overall mean of these estimates $\gbb = \frac{1}{k} \sum_{j=1}^k \gb_j$ provides a point estimator of the optimality gap.  
The sample variance is obtained as in (\ref{eq:var}) by $\varh{\gbb} = \frac{1}{k} \frac{1}{k-1} \sum_{j=1}^k (\gb_j - \gbb)^2$, which results in a one-sided CI $\left[0, \gbb + t_{k-1,\alpha} \sqrt{\varh{\gbb}} \right]$ that has a level of significance which is approximately $\alpha$.

@@

The optimality gap estimator for each batch $\gb_{lj}$ is defined like (\ref{eq:mrp_gap}) for general values of the nonoverlap parameter $\gamma_l$.  
We just removed the minimization in (\ref{eq:mrp_gap}) and used directly an optimal solution of $\xs_j$ of batch $j$.  
The overall mean, $\gbb_l$, is defined a little differently.  
Here, $\gbb_l$ still uses each observation $i = 1, 2, \dots, n_l$ but also makes use of all the information collected throughout the batches.  
That is, if observation $\xiti$ is used in $|B(i)|$ number of batches, all the optimal solutions $\xs_j$ corresponding to each batch $j \in B(i)$ are used for the lower bound estimator.  
Then, $VG_l$ is defined in a similar fashion for the variably overlapping batches variance estimator (\ref{eq:variance_gamma}).

\section{Theoretical Results} 
\label{sec:theory}

@@

First, we restrict our attention to i.i.d.\ sampling by A1.
The classical variance-reduction results for overlapping batches show that $\var{n_l VG_l}$ is reduced by overlapping, thus requiring at least finite fourth moments. 
Finally, our proof relies on an exponential rate of convergence of sampled solution to the true solution of the problem.  
Assumption A3 can be satisfied in several ways.  
Shapiro and Homem-de-Mello \citep{shapiro2000rate} establish rates of convergence for a class of (\ref{eq:sto_prog}), which include two-stage stochastic linear programs with recourse when the support of $\xit$ is finite and the optimal solution is unique.   
Kleywegt, Shapiro,and Homem-de-Mello \citep{kleywegt2002sample} show that assumption A3 can be satisfied for discrete stochastic programs (i.e., when $X$ is a finite set) under the condition of a unique optimal solution $\xs$, and that the moment generating function of $f(\xs,\xit) - f(\x,\xit)$ for $x \in X \backslash \{\xs\}$ is finite valued on $\R$.  
This class of problems includes stochastic integer programs with recourse.
  We present computational results on both stochastic linear and integer programs with recourse in \S \ref{sec:comp}.

@@

The internal optimization in the batches in (\ref{eq:gbar}) makes a straightforward statistical analysis of the behavior of the estimators difficult.  
To overcome this problem, we introduce the following unbiased optimality gap estimators:
\begin{align}
	\db_{lj} & = \frac{1}{m_l} \sum_{i=1}^{m_l} f(\xh,\xit^{\gamma_l(j-1)+i}) - \frac{1}{m_l} \sum_{i=1}^{m_l} f(\xs,\xit^{\gamma_l(j-1)+i}),\ \ \ j = 1, 2, \dots, \nbl, \label{eq:dbar} \\
	\dbb_l & = \frac{1}{n_l} \sum_{i=1}^{n_l} \left[ f(\xh,\xiti) - f(\xs,\xiti) \right], \label{eq:dbb} \\
	VD_l & = \frac{1}{\left( \tfrac{n_l}{m_l} - 1 \right) \left(\nbl\right)} \sum_{j=1}^{\nbl} (\db_{lj} - \dbb_l)^2. \label{eq:vd}
\end{align}
These are essentially the same as the variably overlapping batches estimators in \S \ref{ssec:overlap} with $y^i = f(\xh,\xiti) - f(\xs,\xiti)$.  
These are also defined identically to the original estimators (\ref{eq:gbar})--(\ref{eq:vg}) with the exception that $\xs_j$ from (\ref{eq:gbar}) and (\ref{eq:gbb}) is replaced by an optimal solution $\xs$ in (\ref{eq:dbar}) and (\ref{eq:dbb}).  
With $\xh$ and $\xs$ fixed, estimators (\ref{eq:dbar})--(\ref{eq:vd}) have the same statistical properties as variably overlapping batches estimators in \S \ref{ssec:overlap}.  
Note that the optimal solution $\xs$ is not known. 
However, the estimators (\ref{eq:dbar})--(\ref{eq:vd}) are used only to show convergence of (\ref{eq:gbar})--(\ref{eq:vg}); they are not necessary for carrying out the OMRP algorithm.  
Throughout the rest of this section, we assume that the optimality gap estimators (\ref{eq:gbar})--(\ref{eq:vg}) and their nonoptimized counterparts (\ref{eq:dbar})--(\ref{eq:vd}) use identical samples and values of $n_
l$, 
$m_l$, and $\gamma_l$.  \smallskip

@@

@@

\begin{proof}
	\begin{align}
		\p{\dbb_l - \gbb_l \neq 0} & \leq \p{\left( \bigcap_{j=1}^{\nbl} \{\xs_{j} = \xs\} \right)^C} \notag \\
		& = \p{ \bigcup_{j=1}^{\nbl} \{\xs_{j} \neq \xs\}} \notag \\
% 		& \leq \sum_{j=1}^{\nbl} \p{ \xs_j \neq \xs } \notag \\
		& \leq \left( \nbl\right) \p{ \xs_{m_l} \neq \xs }. \notag
	\end{align}
	With a batching structure in which $m_l = \lfloor n_l^{r} \rfloor$ for some $0<r<1$,  $\gamma_l \geq 1$, and by A3, the desired result follows.
\end{proof}

@@

\begin{theorem} \label{thm:strong_consistency}
	Under assumptions A1--A3, (i) $\gbb_l - \dbb_l \rightarrow 0$, a.s, and (ii) $n_l VG_l - n_l VD_l \rightarrow 0$, a.s. as $l \rightarrow \infty$.  
        Thus, $\gbb_l$ is a consistent estimator of $\mu_{\xh} = \e{\left[f(\xh,\xit) - f(\xs,\xit)\right]}$, and if $n_l VD_l$ is a consistent estimator of $\sigma^2_{\xh} := \var{f(\xh,\xit) - f(\xs,\xit)}$, then $n_l VG_l$ is a consistent estimator of $\sigma^2_{\xh}$.
\end{theorem}

\begin{proof}
	(i) By Lemma \ref{lem:gbb_prob} we saw that $\p{\dbb_l - \gbb_l \neq 0} \rightarrow 0$ exponentially fast.  
        Thus $\sum_l \p{\dbb_l - \gbb_l \neq 0} < \infty$.  
        The desired result follows by the Borel-Cantelli Lemma.
%
(ii) This proof is nearly identical to above.  
Note that
	\[
		\p{n_l VD_l -n_l VG_l \neq 0} \leq \p{\left( \bigcap_{j=1}^{\nbl} \{\xs_{j} = \xs\} \right)^C}
	\]
	and by the proof of Lemma \ref{lem:gbb_prob}, $\p{n_l VD_l -n_l VG_l \neq 0} \rightarrow 0$ exponentially fast, and thus $n_l VG_l$ is a consistent estimator of $\sigma^2_{\xh}$ if $n_l VD_l$ is, by the Borel-Cantelli Lemma.
\end{proof}

Part (i) of Theorem 1 provides conditions under which the point estimator $\gbb$ of OMRP is a consistent estimator of the optimality gap, by establishing almost sure convergence of $\gbb_l - \dbb_l$ to $0$ as $l \rightarrow \infty$. 
Notice that under assumption A1, $\dbb_l \rightarrow \mu_\xh$, a.s. by the strong law of large numbers; as a result, $\gbb \rightarrow \mu_\xh$, a.s.

Strong consistency of the variance estimator $n_l VD_l$ have been studied in \citep{damerdji1994strong} for the cases of nonoverlapping and maximally overlapping batches.  
Under the additional moment assumption
\begin{description}
	\item[A2$^\prime$]  $\exists \epsilon > 0$ such that $\sup_{\x \in X} \e{f(\x,\xi)^{4+\epsilon}} < \infty$,
\end{description}
for a sampling scheme of $m_l = \lfloor n_l^r \rfloor$, Corollaries 3.4 and 4.2 of \citep{damerdji1994strong} show that $n_l VD_l$ is consistent if $\tfrac{2}{4+\epsilon} < r < 1$ for nonoverlapping batches and $\tfrac{2}{4+\epsilon} < r < \tfrac{1}{2}$ for maximally overlapping batches.

>>>>

Taking $n$ observations of the random variables $\xit$ and splitting them into $k$ nonoverlapping batches of size $m$, define 
\begin{align} 
	\gb_j  & = \frac{1}{m} \sum_{i = 1}^{m} f(\xh, \xit^{m(j-1)+i}) - \min_{\x \in X} \frac{1}{m} \sum_{i=1}^{m} f(\x, \xit^{m(j-1)+i}), \ \ \ \ j = 1,2, \dots, k \label{eq:mrp_gap}\\
         & = \frac{1}{m} \sum_{i = 1}^{m} \left[ f(\xh, \xit^{m(j-1)+i}) -  f(\xs_j, \xit^{m(j-1)+i}) \right], \ \ \ \ j = 1,2, \dots, k, \notag
\end{align}
where $\xs_j$ denotes an optimal solution to (\ref{eq:sto_prog_m}) of batch $j$ given on the right-hand side of (\ref{eq:mrp_gap}).    
Since we are assessing the quality of a given solution $\xh \in \X$, we suppress it from the notation $\gb_j$.
The optimality gap estimator $\gb_j$ is similar to $\yb_j$ with individual observations $y^{m(j-1)+i} = f(\xh,\xit^{m(j-1)+i}) - f(\xs_j,\xit^{m(j-1)+i})$, for $i=1,\ldots,m$ and $j=1,\ldots, k$.
Notice that, in this case, $y^i$ not only depends on observation $i$ of batch $j$, $\xit^{m(j-1)+i}$, but all the observations in that batch $\xit^{m(j-1)+1},\xit^{m(j-1)+2},\ldots,\xit^{m(j-1)+m}$ because $\xs_j$ is the optimal solution to (\ref{eq:sto_prog_m}) given in (\ref{eq:mrp_gap}).   
As before, after $k$ estimates are obtained, the overall mean of these estimates $\gbb = \frac{1}{k} \sum_{j=1}^k \gb_j$ provides a point estimator of the optimality gap.  
The sample variance is obtained as in (\ref{eq:var}) by $\varh{\gbb} = \frac{1}{k} \frac{1}{k-1} \sum_{j=1}^k (\gb_j - \gbb)^2$, which results in a one-sided CI $\left[0, \gbb + t_{k-1,\alpha} \sqrt{\varh{\gbb}} \right]$ that has a level of significance which is approximately $\alpha$.

@@

The optimality gap estimator for each batch $\gb_{lj}$ is defined like (\ref{eq:mrp_gap}) for general values of the nonoverlap parameter $\gamma_l$.  
We just removed the minimization in (\ref{eq:mrp_gap}) and used directly an optimal solution of $\xs_j$ of batch $j$.  
The overall mean, $\gbb_l$, is defined a little differently.  
Here, $\gbb_l$ still uses each observation $i = 1, 2, \dots, n_l$ but also makes use of {\it all} the information collected throughout the batches.  
That is, if observation $\xiti$ is used in $|B(i)|$ number of batches, all the optimal solutions $\xs_j$ corresponding to each batch $j \in B(i)$ are used for the lower bound estimator.  
Then, $VG_l$ is defined in a similar fashion for the variably overlapping batches variance estimator (\ref{eq:variance_gamma}).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Theoretical Results} 
\label{sec:theory}

@@

First, we restrict our attention to i.i.d.\ sampling by A1.
The classical variance-reduction results for overlapping batches show that $\var{n_l VG_l}$ is reduced by overlapping, thus requiring at least finite fourth moments. 
Finally, our proof relies on an exponential rate of convergence of sampled solution to the true solution of the problem.  
Assumption A3 can be satisfied for different classes of problems with unique optimal solutions.  
\citet{shapiro2000rate} establish exponential rates of convergence for a class of (\ref{eq:sto_prog}), which include two-stage stochastic linear programs with recourse when the support of $\xit$ is finite.   
\citet{kleywegt2002sample} show that assumption A3 can be satisfied for discrete stochastic programs (i.e., when $X$ is a finite set) under the condition that the moment generating function of $f(\xs,\xit) - f(\x,\xit)$ for $x \in X \backslash \{\xs\}$ is finite valued on $\R$.  
This class of problems includes stochastic integer programs with recourse.
  We present computational results on both stochastic linear and integer programs with recourse in \S \ref{sec:comp}.

@@

The internal optimization in the batches in (\ref{eq:gbar}) makes a straightforward statistical analysis of the behavior of the estimators difficult.  
To overcome this problem, we introduce the following unbiased optimality gap estimators:

\begin{align}
	\db_{lj} & = \frac{1}{m_l} \sum_{i=1}^{m_l} f(\xh,\xit^{\gamma_l(j-1)+i}) - \frac{1}{m_l} \sum_{i=1}^{m_l} f(\xs,\xit^{\gamma_l(j-1)+i}),\ \ \ j = 1, 2, \dots, \nbl, \label{eq:dbar} \\
	\dbb_l & = \frac{1}{n_l} \sum_{i=1}^{n_l} \left[ f(\xh,\xiti) - f(\xs,\xiti) \right], \label{eq:dbb} \\
	VD_l & = \frac{1}{\left( \tfrac{n_l}{m_l} - 1 \right) \left(\nbl\right)} \sum_{j=1}^{\nbl} (\db_{lj} - \dbb_l)^2. \label{eq:vd}
\end{align}
These are essentially the same as the variably overlapping batches estimators in \S \ref{ssec:overlap} with $y^i = f(\xh,\xiti) - f(\xs,\xiti)$.  
These are also defined identically to the original estimators (\ref{eq:gbar})--(\ref{eq:vg}) with the exception that $\xs_j$ from (\ref{eq:gbar}) and (\ref{eq:gbb}) is replaced by an optimal solution $\xs$ in (\ref{eq:dbar}) and (\ref{eq:dbb}).  
With $\xh$ and $\xs$ fixed, estimators (\ref{eq:dbar})--(\ref{eq:vd}) have the same statistical properties as variably overlapping batches estimators in \S \ref{ssec:overlap}.  
Note that the optimal solution $\xs$ is not known. 
However, the estimators (\ref{eq:dbar})--(\ref{eq:vd}) are used only to show convergence of (\ref{eq:gbar})--(\ref{eq:vg}); they are not necessary for carrying out the OMRP algorithm.  
Throughout the rest of this section, we assume that the optimality gap estimators (\ref{eq:gbar})--(\ref{eq:vg}) and their nonoptimized counterparts (\ref{eq:dbar})--(\ref{eq:vd}) use identical samples and values of $n_
l$, 
$m_l$, and $\gamma_l$.  \smallskip

@@

We begin with two lemmas that establish exponential rates of convergence of certain intermediate quantities.

@@

\begin{proof} Let $E^C$ denote the complement of event $E$. Note that
	\begin{align}
		\p{\dbb_l - \gbb_l \neq 0} & \leq \p{\left( \bigcap_{j=1}^{\nbl} \{\xs_{j} = \xs\} \right)^C} \notag \\
		& = \p{ \bigcup_{j=1}^{\nbl} \{\xs_{j} \neq \xs\}} \notag \\
% 		& \leq \sum_{j=1}^{\nbl} \p{ \xs_j \neq \xs } \notag \\
		& \leq \left( \nbl\right) \p{ \xs_{m_l} \neq \xs }. \notag
	\end{align}
	With a batching structure in which $m_l = \lfloor n_l^{r} \rfloor$ for some $0<r<1$,  $\gamma_l \geq 1$, and by A3, the desired result follows.
\end{proof}

@@

We are now ready to show that the resulting point estimators are strongly consistent. 
We use the abbreviation a.s.\ to denote almost sure convergence. 
We also use $\mu_{\xh} := \e{\left[f(\xh,\xit) - f(\xs,\xit)\right]}$ to denote the optimality gap of $\xh$ and $\sigma^2_{\xh} := \var{f(\xh,\xit) - f(\xs,\xit)}$ to denote the associated variance term.

\begin{theorem} \label{thm:strong_consistency}
	Under assumptions A1--A3, (i) $\gbb_l - \dbb_l \rightarrow 0$, a.s., and (ii) $n_l VG_l - n_l VD_l \rightarrow 0$, a.s.\ as $l \rightarrow \infty$.  
        Therefore, $\gbb_l$ is a consistent estimator of $\mu_{\xh}$, and if $n_l VD_l$ is a consistent estimator of $\sigma^2_{\xh}$, then $n_l VG_l$ is a consistent estimator of $\sigma^2_{\xh}$.
\end{theorem}

\begin{proof}
	(i) By Lemma \ref{lem:gbb_prob}, $\p{\dbb_l - \gbb_l \neq 0}$ $\rightarrow 0$ exponentially fast.  
        Thus $\sum_l \p{\dbb_l - \gbb_l \neq 0} < \infty$.  
        The desired result follows by the Borel-Cantelli Lemma.
%
(ii) This proof is nearly identical to above.  
Note that
	\[
		\p{n_l VD_l -n_l VG_l \neq 0} \leq \p{\left( \bigcap_{j=1}^{\nbl} \{\xs_{j} = \xs\} \right)^C}
	\]
	and by the proof of Lemma \ref{lem:gbb_prob}, $\p{n_l VD_l -n_l VG_l \neq 0} \rightarrow 0$ exponentially fast, and thus $n_l VG_l$ is a consistent estimator of $\sigma^2_{\xh}$, if $n_l VD_l$ is, by the Borel-Cantelli Lemma.
\end{proof}

Part (i) of Theorem \ref{thm:strong_consistency} provides conditions under which the point estimator $\gbb$ of OMRP is a consistent estimator of the optimality gap, by establishing almost sure convergence of $\gbb_l - \dbb_l$ to $0$ as $l \rightarrow \infty$. 
Notice that under assumption A1, $\dbb_l \rightarrow \mu_\xh$, a.s. by the strong law of large numbers; as a result, $\gbb \rightarrow \mu_\xh$, a.s. 
Similarly, part (ii) of Theorem \ref{thm:strong_consistency} establishes almost sure convergence of $n_l VG_l - n_l VD_l$ to $0$ as $l \rightarrow \infty$. 
As a result, if $n_l VD_l$ is a consistent estimator of $\sigma^2_{\xh}$, so is $n_l VG_l$.  
Strong consistency of the variance estimator $n_l VD_l$ have been studied in \citep{damerdji1994strong} for the cases of nonoverlapping and maximally overlapping batches.  
Under the additional moment assumption
\begin{description}
	\item[A2$^\prime$]  $\exists \epsilon > 0$ such that $\sup_{\x \in X} \e{f(\x,\xi)^{4+\epsilon}} < \infty$,
\end{description}
for a sampling scheme of $m_l = \lfloor n_l^r \rfloor$, Corollaries 3.4 and 4.2 of \citep{damerdji1994strong} show that $n_l VD_l$ is consistent if $\tfrac{2}{4+\epsilon} < r < 1$ for nonoverlapping batches and $\tfrac{2}{4+\epsilon} < r < \tfrac{1}{2}$ for maximally overlapping batches.
